An exposition on the propriety of restricted Boltzmann machines

### 2016-10-01 writing/appendix-instability.Rmd
fix references and alter last instability proof

\begin{align*}
&\max\left\{\log\frac{P_{\boldsymbol \theta}(v_1, \dots, v_V)}{P_{\boldsymbol \theta}(v^*_1, \dots, v^*_V)}: (v_1, \dots, v_V) \text{ and } (v^*_1, \dots, v^*_V) \in \{-1, 1\}^V \text{ differ by one component} \right\} \\
&\quad = \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
\right)}: v_k \not= v^*_k \right\} \\
& \quad \ge \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{2^H\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
\right)}: v_k \not= v^*_k \right\} \\
&\quad = \max\limits_{k \in 1, \dots, V}\left\{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{
\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\}\right. \\
& \quad \quad \left. - \max\limits_{\boldsymbol h \in \{-1,1\}^H}
\left\{\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
\right\} - H\log2: v_k \not= v^*_k \right\} \\
&\quad \ge \max\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - \min\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - H\log 2
\end{align*}

>>>>

{\bf Proof of (\ref{eqn:lowerbnd}).} With algebra, we have

\begin{align*}
&\log\left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)} \right] = \\
& \qquad \log\left[\frac{\max\limits_{(v_1, \dots, v_V)}\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}
{\min\limits_{(v_1, \dots, v_V)}\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}\right] \\
& \qquad \ge \log\frac{\max\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}
{\min\limits_{(v_1, \dots, v_V)}2^H\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)} \\
& \qquad = \max\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\} - \\
& \qquad \qquad \min\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\} -H\log2 \\
& \qquad \equiv R(\boldsymbol \theta)
\end{align*}
using above the monotonicity of $\log(\cdot)$.

### 2016-10-01 writing/draft.Rmd
cleanup center of universe appendix
fix references and alter last instability proof

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding in section \ref{appendix}.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$. The proof of this is included in appendix \ref{on-the-one-to-one-correspondence-between parameters-and moments-in-the-rbm-distribution}. And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

# Appendix

## On the one-to-one correspondence between parameters and moments in the RBM distribution

@@

## Instability proofs

@@

>>>>

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding and propositions on instability in section \ref{appendix}.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$. The proof of this is included in appendix \ref{center-of-universe}. And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

\begin{appendix}
\section{Appendices}
\label{appendix}
\newtheorem{apptheorem}{Theorem}[section]
\newtheorem{applemma}{Lemma}[section]

\subsection{On the one-to-one correspondence between parameters and moments in the RBM distribution}
\label{center-of-universe}

@@

\subsection{Instability proofs}
\label{instability-proofs}

@@

\end{appendix}
